---
title: "Getting Started with OmicsFlow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with OmicsFlow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

OmicsFlow provides a robust orchestration layer for multi-omics preprocessing, latent feature extraction via MOFA+, and downstream clinical survival forecasting. This vignette outlines the standard user workflow for setting up and executing the pipeline using the R package API.

## 1. Installation

Install the lightweight package core directly from GitHub:

```{r eval=FALSE}
# Install the core orchestration wrapper
remotes::install_github("Abdullah-I-Ali/OmicsFlow")
library(OmicsFlow)
```

Because OmicsFlow orchestrates heavy analytical pipelines (e.g., `MOFA2`, `clusterProfiler`, `xgboost`), we supply a helper function to install the complete scientific stack automatically:

```{r eval=FALSE}
# Installs heavy CRAN and Bioconductor dependencies into the environment
install_omicsflow_dependencies()
```

## 2. Metadata Generation

OmicsFlow uses an abstraction layer to decouple pipeline execution from sample-naming conventions. Before running the pipeline, generate the required metadata templates:

```{r eval=FALSE}
# Provide the omics matrices you possess (any subset is allowed)
res_templates <- generate_metadata_templates(
  rna = "path/to/rna.rds",
  meth = "path/to/meth.rds",
  output_dir = "results/templates"
)
```

This generates `sample_metadata.csv` and `custom_clinical_template.tsv`. You should edit these files to map your samples to their clinical attributes before proceeding.

## 3. Validation

Before kicking off a long multi-omics run, OmicsFlow provides a pre-flight validation checker. This ensures files exist, required schema variables match, and matrix sample IDs perfectly align with the metadata.

```{r eval=FALSE}
res_val <- validate_inputs(
  rna = "path/to/rna.rds",
  meth = "path/to/meth.rds",
  metadata = "results/templates/sample_metadata.csv",
  clinical = "results/templates/custom_clinical_template.tsv",
  clinical_map = "results/templates/clinical_map.json"
)

# Returns a structured object
print(res_val$status)
print(res_val$valid)
```

## 4. Subset Execution

OmicsFlow dynamically supports any combination of RNA, Methylation, CNV, and SNV (minimum 2 modalities required for MOFA+ integration). To run the pipeline on a subset (e.g., RNA and Methylation):

```{r eval=FALSE}
res_flow <- omicsflow(
  rna = "path/to/rna.rds",
  meth = "path/to/meth.rds",
  metadata = "results/templates/sample_metadata.csv",
  clinical = "results/templates/custom_clinical_template.tsv",
  clinical_map = "results/templates/clinical_map.json",
  outdir = "results/my_analysis",
  mofa_iter = 1000,
  render_report = TRUE
)

# Inspect the structured return object
print(res_flow$status)
print(res_flow$executed_modalities)
```

## 5. Report Generation

If `render_report = TRUE` is specified, OmicsFlow will automatically compile a Quarto-driven HTML dashboard upon completion. This report includes:

- Interactive PCA plots
- MOFA+ Latent Factor visualizations
- Machine Learning Survival curves (Kaplan-Meier)
- Pathway Enrichment maps

You can locate the final report in the defined output directory under `reports/OmicsFlow_Report.html`.
