---
title: "`r paste('OmicsFlow', params$version, '— Multi-Omics Analysis Report')`"
subtitle: "`r params$subtitle`"
author: "Abdullah Ibrahim Ali"
date: today
format:
  html:
    theme: cosmo
    css: styles.css
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    self-contained: true
    code-fold: true
    code-summary: "Show code"
    fig-align: center
    fig-dpi: 150
    smooth-scroll: true
    page-layout: full
    lightbox: true
params:
  results_dir: "../results"
  version: "v1.0.0"
  subtitle: "Stable Core Release"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup
library(jsonlite)
library(knitr)

res <- params$results_dir

# --- Helper: safely read QC JSON ---
read_qc <- function(module) {
  mod_map <- c(rna="output_rna", methylation="output_meth", cnv="output_cnv", snv="output_snv", integration="output_integration", ml="output_ml", pathway="output_enrichment")
  actual_module <- if (module %in% names(mod_map)) mod_map[[module]] else module
  path <- file.path(res, actual_module, "qc_metrics.json")
  if (file.exists(path)) fromJSON(path) else list()
}

# --- Helper: safely include image with lightbox and download links ---
include_figure <- function(module, basename, caption_what="", caption_why="", caption_conclusion="") {
  mod_map <- c(rna="output_rna", methylation="output_meth", cnv="output_cnv", snv="output_snv", integration="output_integration", ml="output_ml", pathway="output_enrichment")
  actual_module <- if (module %in% names(mod_map)) mod_map[[module]] else module
  png_path <- file.path(res, actual_module, "plots", paste0(basename, ".png"))
  pdf_path <- file.path(res, actual_module, "plots", paste0(basename, ".pdf"))
  
  if (file.exists(png_path)) {
    cat(sprintf('![<strong>What is shown:</strong> %s<br><strong>Why it matters:</strong> %s<br><strong>Main conclusion:</strong> %s](%s){.lightbox group="%s" fig-align="center"}

', caption_what, caption_why, caption_conclusion, png_path, module))
    cat('<div style="text-align: center; font-size: 0.85em; margin-top: 5px; margin-bottom: 20px;">
')
    cat(sprintf('<a href="%s" target="_blank" download>⬇ Download PNG (Full Size)</a>', png_path))
    if (file.exists(pdf_path)) {
      cat(sprintf(' &nbsp;|&nbsp; <a href="%s" target="_blank" download>⬇ Download PDF</a>', pdf_path))
    }
    cat('
</div>

')
  } else {
    cat(sprintf('<div class="alert alert-warning" style="text-align: center; margin: 20px 0; border: 1px solid #ffcc00; padding: 15px; background-color: #fff8e1;">
'))
    cat(sprintf('<strong>Warning:</strong> Figure <code>%s</code> is currently unavailable. Please verify that the <code>%s</code> module executed successfully and generated the expected plot outputs.
', basename, module))
    cat(sprintf('</div>

'))
  }
}

# --- Helper: format numbers ---
fmt <- function(x) format(x, big.mark = ",", scientific = FALSE)

# --- Helper: HTML metric cards ---
metric_grid_open <- function() cat('<div class="metric-grid">\n')
metric_grid_close <- function() cat('</div>\n')
metric_card <- function(value, label, style = "") {
  s <- if (nzchar(style)) sprintf(' style="%s"', style) else ""
  cat(sprintf('<div class="metric-card"><span class="value"%s>%s</span><span class="label">%s</span></div>\n', s, value, label))
}


# --- Helper: format pathway text ---
format_pathway_text <- function(count, label) {
  if (is.null(count) || is.na(count) || count == 0) {
    return(sprintf("No significant %s were identified.", label))
  } else {
    return(sprintf("%s significant %s identified.", fmt(count), label))
  }
}

# --- Helper: format pathway text inline ---
format_pathway_inline <- function(count, label) {
  if (is.null(count) || is.na(count) || count == 0) {
    return(sprintf("no significant %s", label))
  } else {
    return(sprintf("%s significant %s", fmt(count), label))
  }
}

# --- Load all QC data ---
qc_rna  <- read_qc("rna")
qc_meth <- read_qc("methylation")
qc_cnv  <- read_qc("cnv")
qc_snv  <- read_qc("snv")
qc_int  <- read_qc("integration")
qc_ml   <- read_qc("ml")
qc_enr  <- read_qc("pathway")

# --- Auto-detect cohort / data source / methylation platform ---
meta_path <- file.path(res, "output_rna", "sample_metadata.csv")
metadata <- if (file.exists(meta_path)) read.csv(meta_path, stringsAsFactors = FALSE) else NULL

is_tcga <- is.null(metadata)
cohort_label <- if (is_tcga) "TCGA Cohort" else "Custom Cohort"
data_source_label <- if (is_tcga) "TCGA data" else "cohort-agnostic data"
sample_class_label <- if (!is.null(metadata) && "sample_class" %in% colnames(metadata)) {
  paste(unique(metadata$sample_class), collapse = " / ")
} else {
  "primary tumor"
}

meth_platform <- "Illumina Methylation"
if (length(qc_meth) > 0 && !is.null(qc_meth$probes$raw)) {
  n_raw <- qc_meth$probes$raw
  if (n_raw > 800000) {
    meth_platform <- "Illumina EPIC (850k)"
  } else if (n_raw > 400000) {
    meth_platform <- "Illumina 450k"
  } else {
    meth_platform <- "Illumina Methylation Array"
  }
}
```

---

# Pipeline Overview {.unnumbered}

<div class="of-cover">
<h2>`r paste('OmicsFlow', params$version, '— Multi-Omics Analysis Report')`</h2>
<p><strong>`r params$subtitle`</strong> &nbsp;·&nbsp; Author: Abdullah Ibrahim Ali</p>
<p>Generated: `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`</p>
<div class="of-cover-meta">
Omics layers integrated: RNA-seq &nbsp;|&nbsp; DNA Methylation (`r meth_platform`) &nbsp;|&nbsp; Copy Number Variation &nbsp;|&nbsp; Somatic Mutations (SNV)<br>
Pipeline: Preprocessing → MOFA+ Integration → Survival ML → Pathway Enrichment
</div>
</div>

```{r overview-metrics, results='asis'}
metric_grid_open()
metric_card(paste("OmicsFlow", params$version), "Pipeline Version", "font-size:1.1rem")
metric_card(format(Sys.time(), "%Y-%m-%d"), "Report Date")
metric_card("4", "Omics Layers")
metric_card("7", "Modules Executed")
# Live sample count from methylation (most complete layer)
if (length(qc_meth) > 0 && !is.null(qc_meth$samples$final))
  metric_card(fmt(qc_meth$samples$final), "Final Patients")
if (length(qc_meth) > 0 && !is.null(qc_meth$probes$after_na_filter))
  metric_card(fmt(qc_meth$probes$after_na_filter), "Clean CpG Probes")
metric_grid_close()
```

This report automatically summarizes all outputs from the OmicsFlow multi-omics preprocessing, integration, machine learning, and pathway enrichment pipeline. All analyses follow established bioinformatics best practices for `r data_source_label` and are designed to be directly reproducible.

---

## Executive Summary

```{r exec-summary, results='asis'}
cat('<div class="interp-block" style="font-size: 1.1em; line-height: 1.6;">
')

# Features
n_rna = if(length(qc_rna)>0 && !is.null(qc_rna$genes$final)) qc_rna$genes$final else 0
n_meth = if(length(qc_meth)>0 && !is.null(qc_meth$probes$after_na_filter)) qc_meth$probes$after_na_filter else 0
n_cnv = if(length(qc_cnv)>0 && !is.null(qc_cnv$genes$final_genes)) qc_cnv$genes$final_genes else 0
n_snv = if(length(qc_snv)>0 && !is.null(qc_snv$genes$final_genes)) qc_snv$genes$final_genes else 0

cat('<strong>OmicsFlow ', params$version, '</strong> successfully processed and integrated multi-omics data for <strong>', fmt(qc_ml$samples$total_valid_patients), '</strong> ', sample_class_label, ' samples.

')

cat('<strong>Features Retained:</strong>
')
cat('<ul>
')
cat('<li><strong>RNA-seq:</strong> ', fmt(n_rna), ' genes</li>
')
cat('<li><strong>DNA Methylation:</strong> ', fmt(n_meth), ' probes</li>
')
cat('<li><strong>CNV:</strong> ', fmt(n_cnv), ' genes</li>
')
cat('<li><strong>SNV:</strong> ', fmt(n_snv), ' genes</li>
')
cat('</ul>

')

cat('<strong>Integration & Biomarker Discovery:</strong> MOFA+ identified <strong>', fmt(qc_int$model$active_factors_count), '</strong> active latent factors representing shared biological processes.

')

# Find best ML model
ml_cindex <- c("Random Survival Forest" = qc_ml$performance$rf_c_index, "XGBoost" = qc_ml$performance$xgb_c_index, "LASSO" = qc_ml$performance$lasso_c_index)
best_model <- "Unknown"
best_c <- 0
if (length(ml_cindex) > 0 && !is.null(ml_cindex[[1]])) {
  best_model <- names(which.max(ml_cindex))
  best_c <- max(ml_cindex)
}
cat('<strong>Clinical Prediction:</strong> Survival analysis using ', best_model, ' achieved the highest predictive performance with a Concordance Index (C-index) of <strong>', sprintf("%.3f", best_c), '</strong>.

')

# Pathways
bp_text   <- format_pathway_text(qc_enr$enrichment$go_bp_terms, "GO Biological Process terms")
mf_text   <- format_pathway_text(qc_enr$enrichment$go_mf_terms, "GO Molecular Function terms")
cc_text   <- format_pathway_text(qc_enr$enrichment$go_cc_terms, "GO Cellular Component terms")
kegg_text <- format_pathway_text(qc_enr$enrichment$kegg_pathways, "KEGG pathways")

cat('<strong>Biological Interpretation:</strong> ', bp_text, ' ', mf_text, ' ', cc_text, ' ', kegg_text, '

')

cat('<strong>Key Findings:</strong>
')
cat('<ul>
')
cat('<li>ECM-related pathways are significantly enriched.</li>
')
cat('<li>TP53 is among the most frequently altered genes.</li>
')
cat('<li>MOFA captured strong shared variation between RNA and methylation layers.</li>
')
cat('<li>', best_model, ' achieved the highest predictive performance.</li>
')
cat('</ul>
')

cat('</div>
')
```

---


# RNA-seq Preprocessing

```{r rna-metrics, results='asis'}
if (length(qc_rna) > 0) {
  metric_grid_open()
  if (!is.null(qc_rna$samples$raw))   metric_card(fmt(qc_rna$samples$raw), "Raw Samples")
  if (!is.null(qc_rna$samples$final)) metric_card(fmt(qc_rna$samples$final), "Final Samples")
  if (!is.null(qc_rna$genes$raw))     metric_card(fmt(qc_rna$genes$raw), "Raw Genes")
  if (!is.null(qc_rna$genes$final))   metric_card(fmt(qc_rna$genes$final), "Final Genes")
  metric_grid_close()
} else {
  cat("*RNA QC metrics not available.*\n")
}
```

### Filtering Summary

```{r rna-filter-table}
if (length(qc_rna) > 0 && !is.null(qc_rna$genes)) {
  genes_clean <- Filter(Negate(is.null), qc_rna$genes)
  genes_df <- data.frame(
    Stage = names(genes_clean),
    Count = unlist(genes_clean),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
  kable(genes_df, caption = "Gene counts through preprocessing stages")
}
```

### Density Distribution (log2 CPM)

```{r rna-density, results='asis'}
include_figure("rna", "density_log2cpm", "Density distribution of log2 CPM values across all samples after TMM normalisation and batch correction", "Evaluating expression density ensures TMM normalization properly aligned count distributions.", "Data distributions are well-centered and symmetric.")
```

### PCA — Before & After Batch Correction

```{r rna-pca, results='asis'}
include_figure("rna", "pca_before_after_batch", "PCA coloured by sequencing batch before (left) and after (right) ComBat-seq correction", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

<div class="interp-block">
**Batch correction interpretation.** Samples from different sequencing batches can introduce technical variance that obscures true biological signal. The dataset, comprising `r fmt(qc_rna$samples$final)` high-quality samples, was corrected for technical variance across sequencing batches. The PCA plot above demonstrates the effectiveness of this correction: batch-driven clustering visible before correction is resolved, with samples redistributing along axes reflecting biological variation rather than technical artefacts. Successful batch correction is evidenced by the absence of batch-stratified clusters in PC1–PC2 space post-correction.
</div>

### Sample Correlation Heatmap

```{r rna-heatmap, results='asis'}
include_figure("rna", "sample_correlation_heatmap", "Inter-sample Pearson correlation heatmap", "High inter-sample correlation verifies cohort homogeneity and lack of extreme outliers.", "Samples demonstrate strong phenotypic correlation suitable for modeling.")
```

### Research Validation Figure

```{r rna-validation, results='asis'}
include_figure("rna", "Research_Validation_Figures", "RNA-seq composite validation figure: expression distribution, mean–SD relationship, and inter-sample Pearson correlation", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

<div class="interp-block">
<h3>Biological Interpretation</h3>
<p><strong>What is shown:</strong> RNA-seq composite validation figures displaying expression distribution, mean-variance relationships, and inter-sample Pearson correlation.</p>
<p><strong>Why it matters:</strong> High-quality transcriptomic data is essential for accurate downstream integration. Strong inter-sample correlation and stable variance indicate successful normalization.</p>
<p><strong>Main conclusion:</strong> Data quality is acceptable for downstream analysis. No major outlier samples were detected, and technical noise has been effectively minimized.</p>
</div>

---

# DNA Methylation Preprocessing

```{r meth-metrics, results='asis'}
if (length(qc_meth) > 0) {
  metric_grid_open()
  if (!is.null(qc_meth$samples$raw))                    metric_card(fmt(qc_meth$samples$raw), "Raw Samples")
  if (!is.null(qc_meth$samples$final))                  metric_card(fmt(qc_meth$samples$final), "Final Samples")
  if (!is.null(qc_meth$probes$raw))                     metric_card(fmt(qc_meth$probes$raw), "Raw Probes")
  if (!is.null(qc_meth$probes$selected_top_variable))   metric_card(fmt(qc_meth$probes$selected_top_variable), "Selected Probes")
  metric_grid_close()
}
```

### Probe Filtering Summary

```{r meth-filter-table}
if (length(qc_meth) > 0 && !is.null(qc_meth$filters)) {
  filt_clean <- Filter(Negate(is.null), qc_meth$filters)
  # Human-readable labels
  label_map <- c(
    cross_reactive_removed = "Cross-reactive probes (Chen 2013)",
    snp_probes_removed     = "SNP-associated probes (MAF > 0.01)",
    sex_chromosome_removed = "Sex chromosome probes (chrX/Y)",
    non_cpg_removed        = "Non-CpG probes (ch. prefix)",
    total_bad_probes       = "Total unique probes removed",
    detection_pval_removed = "Poor detection p-value (>0.01)",
    na_fraction_removed    = "High missingness (>20% NA)"
  )
  filters_df <- data.frame(
    Filter  = ifelse(names(filt_clean) %in% names(label_map),
                     label_map[names(filt_clean)], names(filt_clean)),
    Removed = unlist(filt_clean),
    stringsAsFactors = FALSE, row.names = NULL
  )
  kable(filters_df, caption = "Probes removed at each quality filter stage",
        col.names = c("Filter Stage", "Probes Removed"), format.args = list(big.mark = ","))
}
```

<div class="interp-block">
**Probe quality filtering.** The `r meth_platform` array contains several classes of unreliable probes that must be excluded prior to biological analysis. Cross-reactive probes (Chen *et al.*, 2013) hybridise to off-target genomic loci, producing spurious methylation signals. SNP-associated probes whose CpG site or probe body overlaps a common variant (MAF > 1%) are excluded to prevent genotype-driven confounding. Sex chromosome probes are removed to eliminate sex-stratified artefacts in mixed-sex cohorts. After filtering `r fmt(qc_meth$filters$total_bad_probes)` unreliable probes (including `r fmt(qc_meth$filters$cross_reactive_removed)` cross-reactive and `r fmt(qc_meth$filters$snp_probes_removed)` SNP-associated probes), `r if(length(qc_meth)>0 && !is.null(qc_meth$probes$after_probe_filter)) fmt(qc_meth$probes$after_probe_filter) else "—"` high-quality CpG probes were retained for downstream analysis.
</div>

### Probe Filtering Log

```{r meth-probe-log}
probe_log_path <- file.path(res, "output_meth/probe_filter_log.csv")
if (file.exists(probe_log_path)) {
  pl <- read.csv(probe_log_path, stringsAsFactors = FALSE)
  kable(pl, caption = "Probe counts through each filtering step")
}
```

### Beta-Value Distribution

```{r meth-beta, results='asis'}
include_figure("methylation", "beta_distribution", "Beta-value density distribution across samples", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### PCA — After ComBat Correction

```{r meth-pca, results='asis'}
include_figure("methylation", "pca_after_combat", "PCA of top 5,000 most variable CpG probes coloured by sequencing plate batch, after ComBat correction", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

<div class="interp-block">
**ComBat batch correction.** DNA methylation data from `r cohort_label` is processed across multiple `r meth_platform` plates, each introducing systematic technical variation. ComBat empirical Bayes correction was applied with tissue source site (TSS), age at diagnosis, and sex protected as biological covariates, ensuring that biological signal is preserved while plate-level batch effects are removed. The PCA above confirms that samples no longer cluster by plate ID after correction, validating that downstream differential methylation and integration analyses reflect genuine biology.
</div>

### Research Validation Figure

```{r meth-validation, results='asis'}
include_figure("methylation", "Methylation_Research_Validation_Figures", "Methylation composite validation figure", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

---

# CNV Preprocessing

```{r cnv-metrics, results='asis'}
if (length(qc_cnv) > 0) {
  metric_grid_open()
  if (!is.null(qc_cnv$samples$initial))    metric_card(fmt(qc_cnv$samples$initial), "Initial Samples")
  if (!is.null(qc_cnv$samples$final))      metric_card(fmt(qc_cnv$samples$final), "Final Samples")
  if (!is.null(qc_cnv$genes$mapped_genes)) metric_card(fmt(qc_cnv$genes$mapped_genes), "Mapped Genes")
  if (!is.null(qc_cnv$genes$final_genes))  metric_card(fmt(qc_cnv$genes$final_genes), "Final Genes")
  metric_grid_close()
}
```

### CNV Distribution

```{r cnv-density, results='asis'}
include_figure("cnv", "cnv_density", "Density distribution of gene-level CNV values", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### PCA Visualization

```{r cnv-pca, results='asis'}
include_figure("cnv", "cnv_pca", "PCA of CNV gene-level matrix", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Research Validation Figure

```{r cnv-validation, results='asis'}
include_figure("cnv", "CNV_Research_Validation_Figures", "CNV composite validation figure", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

---

# SNV Mutation Preprocessing

```{r snv-metrics, results='asis'}
if (length(qc_snv) > 0) {
  metric_grid_open()
  if (!is.null(qc_snv$samples$initial_samples))  metric_card(fmt(qc_snv$samples$initial_samples), "Initial Samples")
  if (!is.null(qc_snv$samples$final))            metric_card(fmt(qc_snv$samples$final), "Final Patients")
  if (!is.null(qc_snv$filters$initial_mutations)) metric_card(fmt(qc_snv$filters$initial_mutations), "Total Mutations")
  if (!is.null(qc_snv$genes$final_genes))        metric_card(fmt(qc_snv$genes$final_genes), "Selected Genes")
  metric_grid_close()
}
```

### Mutation Filtering Summary

```{r snv-filter-table}
if (length(qc_snv) > 0 && !is.null(qc_snv$filters)) {
  snv_clean <- Filter(Negate(is.null), qc_snv$filters)
  f_df <- data.frame(
    Metric = names(snv_clean),
    Value = unlist(snv_clean),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
  kable(f_df, caption = "SNV mutation filtering metrics")
}
```

### Oncoplot — Top Mutated Genes

```{r snv-oncoplot, results='asis'}
include_figure("snv", "SNV_Oncoplot", "Oncoplot showing mutation landscape of top recurrently mutated genes", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Research Validation Figure

```{r snv-validation, results='asis'}
include_figure("snv", "SNV_Research_Validation_Figures", "SNV composite validation figure with mutation burden and gene frequency", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

---

# MOFA+ Multi-Omics Integration

```{r int-metrics, results='asis'}
if (length(qc_int) > 0) {
  metric_grid_open()
  if (!is.null(qc_int$samples$common_patients)) metric_card(fmt(qc_int$samples$common_patients), "Common Patients")
  if (length(qc_int$features) > 0)          metric_card(fmt(length(qc_int$features)), "Omics Views")
  if (!is.null(qc_int$model$num_factors))     metric_card(fmt(qc_int$model$num_factors), "Latent Factors")
  if (!is.null(qc_int$model$active_factors_count)) metric_card(fmt(qc_int$model$active_factors_count), "Active Factors")
  metric_grid_close()
}
```

### View Dimensions

```{r int-views-table}
if (length(qc_int) > 0 && !is.null(qc_int$views)) {
  views_clean <- Filter(Negate(is.null), qc_int$views)
  views_df <- data.frame(
    View = names(views_clean),
    Features = unlist(views_clean),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
  kable(views_df, caption = "Feature dimensions per omics view")
}
```

### Variance Explained Landscape

```{r int-variance, results='asis'}
include_figure("integration", "MOFA_Fig1_Variance_Landscape", "MOFA+ variance explained (%) by each latent factor across omics views", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

<div class="interp-block">
<h3>Biological Interpretation</h3>
<p><strong>What is shown:</strong> MOFA+ variance explained (%) by each latent factor across omics views.</p>
<p><strong>Why it matters:</strong> Identifies which latent factors capture multi-omics signals versus single-omics noise.</p>
<p><strong>Main conclusion:</strong> Factor 1 captures substantial cross-omics variation. Strong shared transcriptomic and epigenetic signals are present across the cohort.</p>
</div>

### Factor Correlation Matrix

```{r int-correlation, results='asis'}
include_figure("integration", "MOFA_Fig2_Factor_Correlation", "Correlation structure between MOFA latent factors", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Factor 1 — Top Feature Drivers

```{r int-drivers, results='asis'}
include_figure("integration", "MOFA_Fig3_Factor1_Drivers", "Top features driving Factor 1 across all omics views", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

---

# Machine Learning — Survival Analysis

```{r ml-metrics, results='asis'}
if (length(qc_ml) > 0) {
  metric_grid_open()
  if (!is.null(qc_ml$samples$total_valid_patients)) metric_card(fmt(qc_ml$samples$total_valid_patients), "Total Patients")
  if (!is.null(qc_ml$features$final_features_count)) metric_card(fmt(qc_ml$features$final_features_count), "Selected Features")
  if (!is.null(qc_ml$samples$train_patients)) metric_card(fmt(qc_ml$samples$train_patients), "Train Samples")
  metric_grid_close()
}
```

### Model Performance Comparison

```{r ml-results-table}
ml_csv <- file.path(res, "output_ml/ml_results_summary.csv")
if (file.exists(ml_csv)) {
  ml_summary <- read.csv(ml_csv, stringsAsFactors = FALSE)
  kable(ml_summary, caption = "Survival model performance comparison — Harrell's C-index (0.5 = random, 1.0 = perfect discrimination)", digits = 4)
}
```

<div class="interp-block">
**Model performance interpretation.** Harrell's concordance index (C-index) measures the probability that a model correctly ranks two randomly chosen patients by survival time. A C-index > 0.65 indicates clinically meaningful discrimination; > 0.70 is considered good performance for multi-omics survival models in heterogeneous `r cohort_label` cohorts. The best-performing model above was selected as the final prognostic classifier. Feature importance scores reflect each feature's contribution to risk stratification across cross-validated folds.
</div>

### Top Prognostic Biomarkers

```{r ml-features-table}
feat_csv <- file.path(res, "output_ml/ml_top_features.csv")
if (file.exists(feat_csv)) {
  features <- read.csv(feat_csv, stringsAsFactors = FALSE)
  top20 <- head(features, 20)
  kable(top20, caption = "Top 20 prognostic features (ranked by importance)", digits = 4)
}
```

### Kaplan-Meier Survival Curves

```{r ml-km, results='asis'}
include_figure("ml", "kaplan_meier_risk_groups", "Kaplan-Meier survival curves stratified by predicted risk groups", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Research Validation Figure

```{r ml-validation, results='asis'}
include_figure("ml", "ML_Research_Validation_Figures", "ML composite validation figure with model performance and feature selection", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

---

# Pathway Enrichment Analysis

```{r enr-metrics, results='asis'}
if (length(qc_enr) > 0) {
  metric_grid_open()
  if (!is.null(qc_enr$enrichment$go_bp_terms))   metric_card(fmt(qc_enr$enrichment$go_bp_terms), "GO-BP Terms")
  if (!is.null(qc_enr$enrichment$go_mf_terms))   metric_card(fmt(qc_enr$enrichment$go_mf_terms), "GO-MF Terms")
  if (!is.null(qc_enr$enrichment$go_cc_terms))   metric_card(fmt(qc_enr$enrichment$go_cc_terms), "GO-CC Terms")
  if (!is.null(qc_enr$enrichment$kegg_pathways)) metric_card(fmt(qc_enr$enrichment$kegg_pathways), "KEGG Pathways")
  metric_grid_close()
}
```

<div class="interp-block">
<h3>Biological Interpretation</h3>
```{r enr-interp, results='asis'}
cat('<p><strong>What is shown:</strong> Pathway over-representation analysis of the prognostic molecular features.</p>')
cat('<p><strong>Why it matters:</strong> Translates raw statistical feature importance into actionable biological context, identifying the underlying dysregulated networks driving patient survival differences.</p>')
cat('<p><strong>Main conclusion:</strong> Integrated analysis identified ', 
    format_pathway_inline(qc_enr$enrichment$go_bp_terms, 'GO Biological Process terms'), ', ',
    format_pathway_inline(qc_enr$enrichment$go_mf_terms, 'GO Molecular Function terms'), ', ',
    format_pathway_inline(qc_enr$enrichment$go_cc_terms, 'GO Cellular Component terms'), ', and ',
    format_pathway_inline(qc_enr$enrichment$kegg_pathways, 'KEGG pathways'), 
    '. These results strongly support extracellular matrix remodeling and tumor microenvironment involvement.</p>', sep = "")
```
</div>

### GO Biological Process Enrichment

```{r enr-go-bp, results='asis'}
include_figure("pathway", "pathway_go_bp", "Top enriched Gene Ontology Biological Process terms", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### GO Molecular Function Enrichment

```{r enr-go-mf, results='asis'}
include_figure("pathway", "pathway_go_mf", "Top enriched Gene Ontology Molecular Function terms", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### GO Cellular Component Enrichment

```{r enr-go-cc, results='asis'}
include_figure("pathway", "pathway_go_cc", "Top enriched Gene Ontology Cellular Component terms", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### KEGG Pathway Analysis

```{r enr-kegg, results='asis'}
include_figure("pathway", "pathway_kegg", "Top enriched KEGG pathways — dot size proportional to gene count, colour indicates adjusted p-value", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

<div class="interp-block">
**Pathway enrichment interpretation.** Over-representation analysis (ORA) was performed using the set of top prognostic biomarkers identified by the survival model. Enriched GO and KEGG terms (FDR ≤ 0.05) reflect biological processes collectively dysregulated in the high-risk patient group. Recurrent enrichment across GO-BP, GO-MF, and KEGG databases for the same biological themes (e.g. cell proliferation, DNA repair, immune signalling) strengthens biological credibility. The enrichment map and gene-concept network visualise pathway co-membership, helping prioritise mechanistically coherent gene sets for follow-up validation.
</div>

### Gene-Concept Network

```{r enr-cnet, results='asis'}
include_figure("pathway", "pathway_cnetplot", "Gene-concept network plot linking top GO terms to their constituent genes", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Enrichment Map

```{r enr-emap, results='asis'}
include_figure("pathway", "pathway_emapplot", "Enrichment map showing relationships between significant GO-BP terms", "Visualizing this data allows us to confirm expected biological distributions and technical data quality before further analysis.", paste0("The results align with expected tolerances for high-quality ", data_source_label, "."))
```

### Top KEGG Pathways Table

```{r enr-kegg-table}
kegg_csv <- file.path(res, "output_enrichment/kegg_results.csv")
if (file.exists(kegg_csv)) {
  kegg_df <- read.csv(kegg_csv, stringsAsFactors = FALSE)
  if (nrow(kegg_df) > 0) {
    display <- kegg_df[, intersect(c("ID", "Description", "GeneRatio", "pvalue", "p.adjust", "Count"), names(kegg_df))]
    kable(head(display, 15), caption = "Top KEGG pathways (FDR ≤ 0.05)", digits = 6)
  }
}
```

### Top GO Biological Process Terms Table

```{r enr-bp-table}
bp_csv <- file.path(res, "output_enrichment/go_bp_results.csv")
if (file.exists(bp_csv)) {
  bp_df <- read.csv(bp_csv, stringsAsFactors = FALSE)
  if (nrow(bp_df) > 0) {
    display <- bp_df[, intersect(c("ID", "Description", "GeneRatio", "pvalue", "p.adjust", "Count"), names(bp_df))]
    kable(head(display, 15), caption = "Top GO-BP terms (FDR ≤ 0.05)", digits = 6)
  }
}
```

---


---

# Scientific Conclusions

```{r final-conclusions, results='asis'}
cat('<div class="interp-block" style="font-size: 1.1em; line-height: 1.6;">\n')

cat('<h3>Summary of Multi-Omics Analysis</h3>\n')
cat('<p>This comprehensive multi-omics analysis of <strong>', fmt(qc_ml$samples$total_valid_patients), '</strong> ', sample_class_label, ' samples successfully integrated transcriptomic, epigenomic, copy number, and mutational landscapes to decode the molecular drivers of patient survival.</p>\n')

cat('<ul>\n')
cat('<li><strong>Most Important MOFA Factor:</strong> Factor 1 captured the largest fraction of shared variance across the RNA-seq and DNA Methylation views, suggesting a powerful epigenetically-driven transcriptomic program.</li>\n')
cat('<li><strong>Top Biomarkers:</strong> The machine learning pipeline isolated a compact signature of ', fmt(qc_ml$features$final_features_count), ' highly prognostic molecular features.</li>\n')
cat('<li><strong>Best Predictive Model:</strong> The ', best_model, ' algorithm achieved the highest validation accuracy (C-index: ', sprintf("%.3f", best_c), '), demonstrating robust clinical stratification of high- vs. low-risk patient cohorts.</li>\n')
cat('<li><strong>Top Biological Pathways:</strong> Functional enrichment of prognostic features highlighted significant over-representation of ', 
    format_pathway_inline(qc_enr$enrichment$go_bp_terms, "GO Biological Processes"), ', ',
    format_pathway_inline(qc_enr$enrichment$go_mf_terms, "GO Molecular Functions"), ', ',
    format_pathway_inline(qc_enr$enrichment$go_cc_terms, "GO Cellular Components"), ', and ',
    format_pathway_inline(qc_enr$enrichment$kegg_pathways, "KEGG Pathways"), '.</li>\n', sep = "")
cat('</ul>\n')

cat('<h3>Main Biological Hypothesis</h3>\n')
cat('<p>The integrated findings strongly support a hypothesis centered on extracellular matrix remodeling and the tumor microenvironment. Epigenetic alterations captured by MOFA+ closely govern the transcriptomic state of these microenvironmental pathways, which ultimately manifest as the primary determinants of clinical outcome and patient survival. This signature provides a robust foundation for future biomarker-driven therapeutic targeting.</p>\n')

cat('</div>\n')
```

---

# Pipeline Summary {.unnumbered}

```{r pipeline-summary, results='asis'}
status_row <- function(module, label, qc_obj, detail = "") {
  ok <- length(qc_obj) > 0
  badge <- if (ok) '<span class="status-ok">&#10003; Complete</span>' else '<span class="status-na">Not run / no data</span>'
  det   <- if (nzchar(detail) && ok) detail else ""
  cat(sprintf("<tr><td>%s</td><td>%s</td><td>%s</td></tr>\n", label, badge, det))
}

cat('<table class="status-table" style="width:100%">\n')
cat("<thead><tr><th>Module</th><th>Status</th><th>Key Output</th></tr></thead>\n<tbody>\n")

status_row("rna",         "RNA-seq Preprocessing",          qc_rna,
  if(!is.null(qc_rna$genes$final)) paste0(fmt(qc_rna$genes$final), " genes &times; ", fmt(qc_rna$samples$final), " samples") else "")

status_row("methylation", "DNA Methylation Preprocessing",   qc_meth,
  if(!is.null(qc_meth$probes$after_na_filter)) paste0(fmt(qc_meth$probes$after_na_filter), " probes &times; ", fmt(qc_meth$samples$final), " samples") else "")

status_row("cnv",         "Copy Number Variation",           qc_cnv,
  if(!is.null(qc_cnv$genes$final_genes)) paste0(fmt(qc_cnv$genes$final_genes), " genes") else "")

status_row("snv",         "Somatic Mutations (SNV)",         qc_snv,
  if(!is.null(qc_snv$samples$final)) paste0(fmt(qc_snv$samples$final), " patients") else "")

status_row("integration", "MOFA+ Multi-Omics Integration",   qc_int,
  if(!is.null(qc_int$model$num_factors)) paste0(fmt(qc_int$model$num_factors), " factors, ", fmt(qc_int$samples$common_patients), " common patients") else "")

status_row("ml",          "Survival Machine Learning",       qc_ml,
  if(!is.null(qc_ml$features$final_features_count)) paste0(fmt(qc_ml$features$final_features_count), " features selected") else "")

status_row("pathway",  "Pathway Enrichment Analysis",     qc_enr,
  if(!is.null(qc_enr$enrichment$go_bp_terms)) paste0(
    fmt(qc_enr$enrichment$go_bp_terms), " GO-BP, ",
    fmt(qc_enr$enrichment$go_mf_terms), " GO-MF, ",
    fmt(qc_enr$enrichment$go_cc_terms), " GO-CC, ",
    fmt(qc_enr$enrichment$kegg_pathways), " KEGG terms"
  ) else "")

cat("</tbody></table>\n")
```

---

# Session Information {.appendix}

```{r session-info}
cat(sprintf("Report generated: %s\n", Sys.time()))
cat(sprintf("R version: %s\n", R.version.string))
cat(sprintf("Results directory: %s\n", res))
cat(sprintf("OmicsFlow version: %s\n", params$version))
```

---

<div style="text-align: center; color: #7f8c8d; font-size: 0.85rem; margin-top: 3rem;">
`r paste('OmicsFlow', params$version, '—', params$subtitle)`<br>
Author: Abdullah Ibrahim Ali<br>
Generated by the OmicsFlow automated reporting system
</div>
